---
title: "Benefit Transfer"
author: "Natalie Smith"
date: "October 29, 2024"
format: 
  html:
    code-fold: false
    toc: true
    number-sections: true
    embed-resources: true
theme: Litera
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

# Introduction

Coastal wetlands are essential in providing protection against storm surges and mitigating the effects of sea-level rise. These ecosystems function as natural sponges, absorbing wave energy and significantly reducing flood risk. This study will evaluate the economic value of restoring 60 hectares of salt marsh in Huntington Beach, California, employing a benefits transfer analysis.

Data from Bayraktov et al. (2015) will be incorporated, which includes a meta-analysis of the restoration costs associated with coastal wetlands. Furthermore, insights from Costanza et al. (2021) will be utilized, as this research estimates the storm protection benefits of wetlands through calculations of avoided damages.

# Part 1: Restoration Costs

*Employ a benefit transfer to find the costs of restoring 60 hectares of wetland.*

```{r}
#load libraries
library(tidyverse)
library(here)
library(knitr)
library(kableExtra)

#import data
cost_df <- read.csv(here("data", "cost_df.csv"))
```

## 

*Inspect the dataset. What wetland type is most appropriate for your analysis? Filter the dataset for that wetland type.*

```{r}
#inspect the dataset
#unique(cost_df$wetland_type)

#filter dataset for wetlands type
wetland_df <- cost_df %>% 
  filter(wetland_type == "saltmarsh") %>% 
  drop_na()
```

Saltmarsh is the most appropriate for our analysis.

## 

*Create a new variable with the restoration costs per hectare. Then, choose one of the research studies and estimate the total costs of restoring 60 hectares in Huntington Beach. Explain your choice.*

```{r}
#create a new variable with restoration costs per hectare
wetland_df <- wetland_df %>%
  mutate(cost_per_ha = total_cost_2010 / area_ha)

#show me the study with the closest area to 60 hectares
# wetland_df %>% 
#   filter(area_ha > 40 & area_ha < 100) %>% 
#   select(observation, area_ha, total_cost_2010, cost_per_ha)
```

Given that the salt marsh we aim to restore spans 60 hectares, I will select a study from the dataset with the area that most closely matching this size that is in the United States. The study with the closest match is the one with an observation ID of 126, which covers 44.5 hectares and has a cost of \$213,689.9 per hectare. Based on this rate, the estimated total cost to restore 60 hectares of salt marsh in Huntington Beach is approximately \$12,821,394.

## 

*We are interested in the marginal cost of each additional hectare restored. Make a scatter plot of hectares on the y-axis and total restoration costs on the x-axis in the USA. Describe the relationship between total costs and wetland area restored.*

```{r, fig.cap="Figure 1: Scatterplot illustrating the relationship between total restoration costs and salt marsh area restored, using log-transformed variables."}

#filter wetland df to only include USA
wetland_usa <- wetland_df %>% 
  filter(country=="USA")

# Scatter plot with log-transformed axes
ggplot(wetland_usa, aes(x = total_cost_2010, y =area_ha)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +  # Log transform x-axis
  scale_y_log10() +  # Log transform y-axis
  labs(title = "Total Restoration Costs by Saltmarsh Area",
       x = "Saltmarsh Area (ha, log scale)",
       y = "Total Restoration Costs (USD2010, log scale)") +
  theme_bw()
```

The scatterplot, which features log-transformed variables to better visualize the trends, illustrates a positive trend between total restoration costs and the area of saltmarsh restored, though the relationship isn’t strictly linear. The costs fluctuate suggesting that the marginal cost per additional hectare isn't constant.

## 

*Run the following regression: total_cost_2010 = a + b\*area_ha + error. Using the outcome of this regression, recalculate the total cost of restoring 60 hectares of wetland.*

```{r}
# run regression
lm_cost <- lm(total_cost_2010 ~ area_ha, data = wetland_usa)

# display results
summary(lm_cost)
```

```{r}
# recalculate total cost with 60 ha of wetland
hb_cost <- predict(lm_cost, data.frame(area_ha = 60))
# hb_cost
```

Based on the regression analysis, the estimated total cost to restore 60 hectares of salt marsh in Huntington Beach is approximately \$3,077,755.

## 

*Describe one way you could improve your cost analysis and valuation.*

The cost analysis could be improved by incorporating additional variables that could affect restoration costs per hectare such as project size, sector, site-specific conditions, ecological complexity, and project duration. Incorporating these factors into the regression model would allow for a more accurate estimate of the total cost of restoring 60 hectares of wetland.

# Part 2: Storm Protection Benefits

Costanza et al. (2021) analyzed 1288 coastal storms globally to calculate the storm protection benefits from wetlands. The authors obtained the following regression estimates:

$$
ln(damages/GDP) = -7.992 - 0.236ln(wetlands) + 3.298ln(wind speed) - 
0.55ln(speed) + 0.137(volume) - 0.058(time)
$$

Variables include:

-   Damages/GDP wind speed of the storm (windspeed)
-   Forward speed of the storm (speed)
-   Wetland area in the swath of the storm (wetlands)
-   Volume of water in the ocean proximal to the storm landfall (volume)
-   Year of the storm minus 1900 (time) as a (non-transformed) linear variable.

## 

*Interpret the coefficient on ln(wetlands)*

The coefficient for ln(wetlands) is -0.236, indicating that a 1% increase in wetland area is associated with a 0.236% decrease in damages relative to GDP. Because both the dependent variable (damages/GDP) and the wetland area are log-transformed, this coefficient captures the percentage change in damages/GDP resulting from a 1% change in wetland area. Essentially, larger wetland areas provide greater storm protection benefits, effectively mitigating the economic damages incurred during storms.

## 

*Calculate the avoided damage of 60 additional hectares of wetlands in case of a storm like Hurricane Hilary. You have the following information: damages = \$18 million, and the available wetland area today is 72 hectares. Assume the GDP doesn't change (only damage moves), and all the remaining variables remain constant.*

```{r}

#ln(damages/GDP) = -7.992 - 0.236ln(wetlands) + 3.298ln(wind speed) - 
#0.55ln(speed) + 0.137(volume) - 0.058(time)

#calculate change in wetland 
current_wetland_area <- 72
additional_wetland_area <- 60
total_wetland_area <- current_wetland_area + additional_wetland_area 
# 72 + 60 = 132


ln_wetland <- log(total_wetland_area/current_wetland_area)
# ln_wetland
#log(132/72) = 0.6061358034

#reduction in damages
ln_damages <- -0.236 * ln_wetland
# - 0.236 * 0.6061358034 = -0.1430480496
# ln_damages

#calculate damage
damages <- 18000000  
damage_new <- damages * exp(ln_damages)
#18,000,000 * exp(-0.1430480496)
# damage_new

avoided_damage = damages - damage_new
#18000000 - 15600824 = 2399176
avoided_damage #million

```

Restoring additional wetlands can effectively reduce economic storm damage by approximately \$2,399,176 in events similar to Hurricane Hilary.
